Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-11 paper/ms.tex
ms: improve magnetic B star connection

\begin{figure}[!t]
	\begin{center}
		\centering
		%\includegraphics[width=0.9\textwidth]{f8.pdf}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f8a.pdf}
		}

		\vspace{-0.4cm}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f8b.pdf}
		}
		\vspace{-0.1cm}
		\caption{
      {\bf The magnetic B star connection.}  HD~37776 and HD~64740 (left column) are
      magnetic B stars with field geometries dominated by high order
      multipoles.
			Here, we compare their phased TESS light curves against the
			``best-matching'' CQV light curves (right column), selected by eye from
			Figure~\ref{fig:cqvs}.
      Stellar masses rounded to one significant figure are given
      in the lower right of each panel; the star, TESS sector, and
      period are listed in the title.
		}
		\label{fig:bstar}
	\end{center}
\end{figure}

@@

An argument against the connection between CQVs and the
$\sigma$~Ori~E analogs is that the light curve of $\sigma$~Ori~E is
much simpler than those in Figure~\ref{fig:cqvs}, with only two broad
local minima, and one ``hump'' \citep{2005ApJ...630L..81T, 2022ApJ...924L..10J}.  
Within the model proposed by Townsend et al., this is explained by the simplicity of 
the star's magnetic field, which to first-order is a
dipole, similar to most other magnetic B stars
\citep{2009ARA&A..47..333D}.  This approximation matches the qualitative
behavior of both the broadband light curves, and the
line-profile variations seen in hydrogen, helium, and carbon
\citep{2012MNRAS.419..959O}.

Two exceptions to the rule of ``simple field topologies for magnetic
B stars'' are HD~37776 and HD~64740, which are B stars that were
previously known from spectropolarimetry in the 2000s-2010s to have
field geometries dominated by high order multipoles
\citep{2011ApJ...726...24K}.  In a review of the literature, we
learned that recent TESS light curves of these two B stars appear
surprisingly similar to the CQV light curves
\citep{2020pase.conf...46M}.  Figure~\ref{fig:bstar} shows the phased
TESS light curves for the two outlying magnetic B stars, with by-eye
``best matching'' CQVs shown for comparison.  The number of dips per
cycle, the shapes of the dips, and the dip depths relative to the
sinusoidal envelope, all seem to be quite similar.  The
implication of such a surprising connection would be that the
highly-structured M dwarf light curves imply multipolar magnetic
fields -- since it is the complex fields which are the defining
characteristic of these two B stars, relative to other known magnetic
B stars like $\sigma$~Ori~E.  
This is consistent with the non-axisymmetric topologies that have been
reported in a small number of M dwarfs for which Zeeman Doppler
Imaging is technically feasible 
\citep[see the review by][]{2021A&ARv..29....1K}.

The physical similarity between the B stars and the M dwarfs 
presumably has its origin in the existence of a ``centrifugal magnetosphere'' \citep[see][]{2013MNRAS.429..398P}.
In other
words, both classes of objects probably satisfy the condition $R_{\rm A} > R_{\rm cr}$, for
$R_{\rm A}$ the Alfv\'en radius.
When this condition holds, stable points exist wherever the outward
centrifugal force is balanced by magnetic tension of closed field loops.
In the converse ``dynamical'' case, when  $R_{\rm A} < R_{\rm cr}$, any
material launched by the star returns to its surface over the free-fall timescale.
% While there are some qualitative connections to Jupiter-Io, it's not perfect.
% In that system, R_A>R_cr...
% But the plasma torus exists near Io's 42 hour orbit
% ...and is nonetheless STILL FORCED into corotation at a 10hr period (presumably by Jupiter's B-field).
% This type of structure is very different, because here we actually see the plasma at corotation!

@@

In this work, we searched 120-second cadence TESS data collected
between July~2018 and Sep~2022 for complex quasiperiodic variables
(CQVs).  The target stars were \nstarssearched\ K and M dwarfs
within 150\,pc.  The selection function was $\gtrsim 80\%$ complete within 30\,pc,
and $\lesssim 10\%$ complete at distances exceeding 100\,pc.

In this sample of stars, we found \ngoods\ objects that showed complex
quasiperiodic behavior over at least one TESS sector.  Because the
TESS 120-second stellar sample was rooted in a heterogeneous selection
function that may have been biased in favor of young stars over field
stars, we caution against interpreting our detection fraction in
terms of broader population statistics.  The \ngoods\ bonafide
CQVs are listed in Table~\ref{tab:thetable}.  This table also includes
an additional \nmaybes\ ambiguous CQVs, whose designation is less
certain.

Analyzing the TESS light curves and stellar properties of these
objects, we draw the following conclusions.

\begin{enumerate}[leftmargin=*]
	%
	\item CQVs are quasiperiodic.  The mean periods remain fixed over the
		$>$1{,}000-cycle baseline of available observations; but the light curve shapes
		always evolve (Figure~\ref{fig:evoln}).
  %
  \item The same CQV can show multiple periods simultaneously.
     LP 12-502, for instance, showed dips with four
    distinct periods within $\pm 0.3\%$ of its fundamental period,
    sometimes simultaneously, and each lasting for up to 50 cycles
    (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
    ($>$100\,cycles) and impulsive ($<$1 cycle).  ``State-switches''
    can cause dips to collapse instantaneously, and are often but not
    exclusively linked with observed optical flares.  Dip growth
    however seems to happen over durations of at least ten cycles, and
    slow dip decay can also occur.
  %
    \item The rate of dip evolution can be used to place a model-dependent
    constraint on how much material is either being accreted or ejected
    during the state changes (Section~\ref{subsec:massflux}).
    Order of magnitude estimates require at least an asteroid belt's worth
    of dust ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or 100 times more 
    material if the occulting matter is gas.
  %
  \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
    the existence of multiple CQVs in AB Dor, the Pleiades, and
    Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
    200\,Myr, based on the one CQV we found in the Carina Near
    moving group (TIC~294328887; $\approx$200\,Myr).  The lack of
    detected CQVs in the Hyades and Praesepe suggests that the
    lifetime of the phenomenon is likely limited to the first few
    hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based on
    the fraction of bonafide CQVs that either turned on or turned off during
    TESS re-observations, two years after the initial observation.
  %
    \item While the majority of detected CQVs have masses of
    0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
    might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
    TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
    $\approx$0.6\,$M_\odot$, are the two massive outliers
    (Section~\ref{subsec:extreme}), and we encourage additional
    scrutiny of these objects in future work.
  %
    \item Surprising analogs to the M  and K dwarf CQVs exist in
    magnetic B stars, but only those with multipolar field topologies.
    Since most magnetic B stars have dipolar magnetic fields,
    this suggests that the CQV dips and warps are similarly being sculpted
    by the stellar magnetic fields, and that the magnetic
    fields themselves are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc, and
    the brightest have $V$$\approx$12 ($J$$\approx$7.5).  This
    conclusion is an extrapolation from the CQV sample derived in this work
    being $\gtrsim$80\% complete within 30\,pc.  Expanding our analysis of
    the TESS data to the full frame images would yield a truly
    volume-limited selection function, and would expand the CQV census
    by about a factor of two within 50\,pc, and by a factor of ten
    within 100\,pc.
\end{enumerate}

@@

%TODO: nix this paragraph?  This edge on hypothesis only holds
%TODO if the magnetic field is aligned with the stellar spin axis.
{\bf todo: nix paragraph? technically this only works if B is aligned
with stellar spin axis... which would be aligned w planetary system}
In both the gas and dust scenarios, CQVs are preferentially viewed
edge-on.  This implies that after correcting for the line-of-sight
inclination, at least one third of low mass stars
\citep{2022AJ....163..144G} trap circumstellar material in the same
way.  This suggests that CQVs should preferentially show edge-on
transiting planets, at larger distances from the star than the
corotating material.  Given these points, important follow-up
observations should include searching for outer transiting planets,
and measuring equatorial velocities in order to test whether the
stellar inclination angles are indeed preferentially edge-on.

>>>>

\begin{figure}[!th]
	\begin{center}
		\centering
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f8a.pdf}
		}

		\vspace{-0.4cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f8b.pdf}
		}

		\vspace{-0.4cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f8c.pdf}
		}

		\vspace{-0.1cm}
		\caption{
      {\bf The magnetic B star connection.}
      $\sigma$~Ori~E and HD~345439 (top row) are magnetic B stars with
      predominantly dipolar magnetic fields known to host
      circumstellar plasma tori.  HD~37776 and HD~64740 (middle row)
      are analogous magnetic B stars with field toplogies dominated
      by high order multipoles.  The bottom row compares the latter
      systems against the ``best-matching'' CQV light curves, selected
      by eye from Figure~\ref{fig:cqvs}.  CQVs have light curves that are
      visually similar to the topologically complex magnetic B stars.
      Stellar masses rounded to one significant figure are given in
      the lower right of each panel; the star, TESS sector, and period
      are listed in each subtitle.
		}
		\label{fig:bstar}
	\end{center}
\end{figure}

@@

An argument against the connection between CQVs and the $\sigma$~Ori~E
analogs is that the light curve of $\sigma$~Ori~E is much simpler than
those in Figure~\ref{fig:cqvs}, with only two broad local minima, and
one ``hump'' \citep{2005ApJ...630L..81T, 2022ApJ...924L..10J}.  Within
the model proposed by Townsend et al., this is explained by the
simplicity of the star's magnetic field, which to first-order is a
dipole, similar to most other magnetic B stars
\citep{2007A&A...475.1053A,2009ARA&A..47..333D}.  This approximation
matches the qualitative behavior of both the broadband light curves,
and the line-profile variations seen in hydrogen, helium, and carbon
\citep{2012MNRAS.419..959O}.

Two exceptions to the rule of ``simple field topologies for magnetic
B stars'' are HD~37776 and HD~64740, which are B stars that were
previously known from spectropolarimetry to have
field geometries dominated by high order multipoles
\citep{2011ApJ...726...24K}.  Reviewing the literature, we
learned that recent TESS light curves of these two B stars appear
surprisingly similar to the CQV light curves
\citep{2020pase.conf...46M}.  The middle row of Figure~\ref{fig:bstar}
shows the phased TESS light curves for these two outlying magnetic B
stars, with by-eye ``best matching'' CQVs shown for comparison.  The
number of dips per cycle, the shapes of the dips, and the dip depths
relative to the sinusoidal envelope, all seem to be quite similar.
The implication of this surprising connection is that the
highly-structured M dwarf light curves may imply multipolar magnetic
fields -- since it is the complex fields which are the defining
characteristic of these two B stars, relative to other known magnetic
B stars like $\sigma$~Ori~E.  This is consistent with the
non-axisymmetric topologies that have been reported in a small number
of M dwarfs for which Zeeman Doppler Imaging is technically feasible
\citep[see the review by][]{2021A&ARv..29....1K}.

The physical similarity between the B stars and the M dwarfs
presumably has its origin in the existence of a ``centrifugal
magnetosphere'' \citep[see][]{2013MNRAS.429..398P}.  In other words,
both classes of objects probably satisfy the condition $R_{\rm A} >
R_{\rm cr}$, for $R_{\rm A}$ the Alfv\'en radius.  When this condition
holds, stable points exist wherever the outward centrifugal force is
balanced by magnetic tension of closed field loops.  In the converse
``dynamical'' case, when  $R_{\rm A} < R_{\rm cr}$, any material
launched by the star returns to its surface over the free-fall
timescale.

% While there are some qualitative connections to Jupiter-Io, it's not
% perfect.  In that system, R_A>R_cr...  But the plasma torus exists
% near Io's 42 hour orbit ...and is nonetheless STILL FORCED into
% corotation at a 10hr period (presumably by Jupiter's B-field).  This
% type of structure is very different, because here we actually see
% the plasma at corotation!

@@

In this work, we searched 120-second cadence TESS data collected
between July~2018 and Sep~2022 for complex quasiperiodic variables
(CQVs).  The target stars were \nstarssearched\ K and M dwarfs
within 150\,pc.  The selection function was $>$$80\%$ complete within 30\,pc,
and $<$$10\%$ complete at distances exceeding 100\,pc.

In our target sample, we found \ngoods\ objects that showed complex
quasiperiodic behavior over at least one TESS sector.  
These \ngoods\ bonafide
CQVs are listed in Table~\ref{tab:thetable}.  This table also includes
an additional \nmaybes\ ambiguous CQVs, whose designation is less
certain.
We inferred ages for all but two of the \ncpvsfound\ objects based on association
memberships; we also derived temperatures and radii using SED fitting,
and then inferred stellar masses by
interpolating against stellar evolutionary models.
We caution that our selection function was not volume-limited:
the
TESS 120-second stellar sample had a heterogeneous selection
function which may have been biased in favor of young stars over field
stars.
Previous work however has shown that $\approx$1-3\% of M dwarfs
younger than $\approx$100\,Myr show the CQV phenomenon
\citep{2016AJ....152..114R,2022AJ....163..144G}.

Analyzing the TESS light curves and stellar properties of our
CQVs, we draw the following conclusions.

\begin{enumerate}[leftmargin=*]
	%
	\item CQVs are quasiperiodic.  The mean periods remain fixed over the
		$>$1{,}000-cycle baseline of available observations; but the light curve shapes
		always evolve (Figure~\ref{fig:evoln}).
  %
  \item The same CQV can show multiple periods simultaneously.
     LP 12-502, for instance, showed dips with four
    distinct periods within $\pm 0.3\%$ of its fundamental period,
    sometimes simultaneously, and each lasting for up to 50 cycles
    (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
    ($>$100\,cycles) and impulsive ($<$1 cycle).  
    Dip growth seems to happen over durations of at least ten cycles, and slow dip decay can also occur.
    ``State-switches''
    correspond to dips to collapsing instantaneously, and are almost always
    linked with observed optical flares.
    This suggests that the occulting material is responding to sudden changes
    in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a model-dependent
    constraint on how much material is either being accreted or ejected
    during the state changes (Section~\ref{subsec:massflux}).
    Order of magnitude estimates require at least an asteroid belt's worth
    of dust ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or 100 times more 
    material if the occulting matter is gas.
  %
  \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
    the existence of multiple CQVs in AB Dor, the Pleiades, and
    Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
    200\,Myr, based on the one CQV we found in the Carina Near
    moving group (TIC~294328887; $\approx$200\,Myr).  The lack of
    detected CQVs in the Hyades and Praesepe suggests that the
    lifetime of the phenomenon is likely limited to the first few
    hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based on
    the fraction of bonafide CQVs that either turned on or turned off during
    TESS re-observations, two years after the initial observation.
  %
    \item While the majority of detected CQVs have masses of
    0.1-0.5\,$M_\odot$, the upper mass limit for the CQV phenomenon
    might extend into the K dwarf regime ($<$$0.8$\,$M_\odot$).
    TIC~405754448 and TIC~405910546, with masses of $\approx$0.8 and
    $\approx$0.6\,$M_\odot$, are the two massive outliers
    (Section~\ref{subsec:extreme}), and we encourage additional
    scrutiny of these objects in future work.
  %
    \item Surprising analogs to the M  and K dwarf CQVs exist in
    magnetic B stars, but only those with multipolar field topologies.
    Since most magnetic B stars have dipolar magnetic fields,
    this suggests that the CQV dips and warps are similarly being sculpted
    by the stellar magnetic fields, and that the magnetic
    fields themselves are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc, and
    the brightest have $V$$\approx$12 ($J$$\approx$7.5).  This
    conclusion is an extrapolation from the CQV sample derived in this work
    being $\gtrsim$80\% complete within 30\,pc.  Expanding our analysis of
    the TESS data to the full frame images would yield a truly
    volume-limited selection function, and would expand the CQV census
    by about a factor of two within 50\,pc, and by a factor of ten
    within 100\,pc.
\end{enumerate}

@@

In both the gas and dust scenarios, CQVs are preferentially viewed
edge-on.  This implies that after correcting for the line-of-sight
inclination, roughly one third of low mass stars
\citep{2022AJ....163..144G} could trap circumstellar material in the same
way.  It also suggests that CQVs may preferentially show
transiting planets at larger distances than the corotating material, though
this may be dependent on whether the magnetic and stellar
spin axes tend to be aligned.  Given these points, observational follow-up
work should include searching for outer transiting planets,
and measuring equatorial velocities in order to test whether the
stellar inclination angles are indeed preferentially edge-on.
